Comparing Generative Pre-trained Transformer Models to Physicians for Assigning Causes of Death from Verbal Autopsy: A Case Study of 6939 Deaths in Sierra Leone from 2019-2022

### 2024-04-13 manuscript/manuscript.tex
Clean up plots with overlapping labels

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
%%%%

\lstset{
    texcl=true,
    basicstyle=\small\sf,
    commentstyle=\small\rm,
    mathescape=true,
    escapeinside={(*}{*)}
}

@@

Recently, Large Language Models (LLM), leveraging massive datasets and deep learning approaches, have made advances in performing a variety of Natural Language Processing (NLP) tasks using free-text, such as question answering, code generation, and even medical diagnosis \cite{changSurveyEvaluationLarge2023,lundChattingChatGPTHow2023,svyatkovskiyIntelliCodeComposeCode2020,hauptAIGeneratedMedicalAdvice2023}. On November 30, 2022, a widely-available LLM called ChatGPT was released by OpenAI with capabilities of answering natural language text inquiries using training data up to September 2021. ChatGPT-3 was based on several Generative Pre-trained Transformer (GPT) models between 2018 to 2020, namely GPT-1 to GPT-3, which had notable differences in training data sizes of 5 gigabytes to 45 terabytes from web sources that resulted in 117 million to 175 billion parameter models \cite{wuBriefOverviewChatGPT2023}. On March 14, 2023, ChatGPT-4 was released with human-level performance on various professional and academic exams and benchmarks that outperformed ChatGPT-3 \cite{openaiGPT4TechnicalReport2023a}. Given the limited usage of free-text open narratives in computer models for determining CODs, and recent advances in LLMs that leverage natural language text prompts, we conduct a case study with Sierra Leone deaths from VA in 2019 to 2022 to compare four models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, to physicians for determining CODs.

@@

Four computer models were used to assign COD for each of the 6939 physician agreed records: GPT-3.5, GPT-4, InterVA-5 , and InSilicoVA. Each model required pre-processing of the 6939 records into input data, and standardization of output COD codes from models for performance evaluation as not all models produced comparable codes across outputs. Although each model can assign multiple CODs per record, only the first generated COD response from GPT-3.5 and GPT-4, and the most probable COD from InterVA-5 and InSilicoVA were used for evaluation. All model outputs were converted to CGHR-10 codes to evaluate performances of models for COD assignment relative to physicians.

@@

GPT-3.5 \cite{brownLanguageModelsAre2020} and GPT-4 \cite{openaiGPT4TechnicalReport2023a} are LLMs that utilize deep neural networks with transformer architectures \cite{vaswaniAttentionAllYou2017} and reinforcement learning from human feedback \cite{ouyangTrainingLanguageModels2022,christianoDeepReinforcementLearning2017,stiennonLearningSummarizeHuman2020,wirthSurveyPreferencebasedReinforcement2017} to follow instructions from prompts and provide human-level responses, with known differences in GPT-4 possessing multimodal capabilites (e.g. image/voice input/output), more recent training data, and improved responses compared to ChatGPT-3 \cite{wuBriefOverviewChatGPT2023}. GPT-3.5 and GPT-4 were instructed with text prompts to assign CODs based on the open narrative. InterVA-5 and InSilicoVA are widely used and studied standard statistical models \cite{murrayUsingVerbalAutopsy2014,otiVerbalAutopsyInterpretation2010,benaraEvaluationMethodsAssigning2023,chandramohanEstimatingCausesDeath2021,tungaVerbalAutopsyModels2021,leitaoComparisonPhysiciancertifiedVerbal2014,jhaAutomatedPhysicianAssignment2019} for COD assignment in VAs under the openVA framework \cite{liOpenVAToolkitVerbal2023}. InterVA-5 applies Bayesian probabilistic modelling \cite{bayesEssaySolvingProblem1958} using a set of standardized symptoms from reports and related conditional probabilities from medical experts to assign CODs based on the highest probability \cite{byassIntegratedApproachProcessing2019,byassStrengtheningStandardisedInterpretation2012}. InSilicoVA improves upon InterVA (e.g. comparable probabilities across individuals, measures of uncertainty, and inclusion of additional data sources) with a hierarchical Bayesian framework and Markov Chain Monte Carlo (MCMC) simulations \cite{brooksMarkovChainMonte1998,chibMarkovChainMonte2001,hanMarkovChainMonte2001} to incorporate multiple sources of uncertainty for assigning CODs based on the highest probability \cite{mccormickProbabilisticCauseofDeathAssignment2016b}. For assigning CODs, GPT-3.5 and GPT-4 require prompts containing conversation-like textual instructions as input, while InterVA-5 and InSilicoVA require structured symptom and associated sociodemographic data as input.

@@

For GPT-3.5 and GPT-4, 6939 and 6935 text prompts were generated for each physician agreed record respectively, as input to instruct the models to assign CODs based on the open narratives. Two types of text prompts were used: user prompts and system prompts. User prompts were generated with a template involving replacing the placeholder \verb|<narrative>| using the free-text open narratives:

@@

The GPT-3.5 and GPT-4 Application Programming Interface (API) was accessed using Python version 3.11.4 and used to assign CODs for each record. GPT-3.5 used the \verb|gpt-3.5-turbo| model, while GPT-4 used the \verb|gpt-4-0613| mode. The parameter \verb|top_p| for GPT-3.5 and GPT-4, representing the sampling temperature, was set to 0.3 to consider only the top 30\% probability mass from the token results \cite{openaiOpenAIPlatformAPI2024}. Higher values closer to 1 may make the results more random, while lower values closer to 0 make the results more focused.

@@

Of the 6939 input records, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA were able to assign CODs for 6929, 6909, 6805, 6729 records respectively. All 6805 InterVA-5 and 6729 InSilicoVA records with WHO VA 2016 v1.5 output codes \cite{worldhealthorganizationVerbalAutopsyStandards2016} were converted into ICD-10 codes. After all model outputs used ICD-10 codes, they were further converted to CGHR-10 codes. The 6929 GPT-3.5 and 6909 GPT-4 output records with ICD-10 codes were converted into 6892 and 6856 records with CGHR-10 codes, where 37 and 53 records did not have matching CGHR-10 codes respectively. All 6805 InterVA-5 and 6729 InSilicoVA records with ICD-10 codes were converted into records with CGHR-10 codes.

@@

Of the 6939 records, GPT-4 had the highest PCCC followed by GPT-3.5, InSilicoVA, and InterVA-5 at 0.71, 0.64, 0.44, and 0.43 PCCC respectively (Table \ref{tab-perf-overall}). GPT-3.5 and GPT-4 had improvements ranging from 0.20-0.28 PCCC over InSilicoVA and InterVA-5, while GPT-4 slightly improved over GPT-3.5 by 0.07 PCCC. CSMF accuracies were above 0.7 for all models at 0.89, 0.86, 0.74, and 0.78 for GPT-4, GPT-3.5, InSilicoVA, and InterVA-5 respectively. Figure \ref{fig-perf-agegroup} shows the PCCC performance across age groups with reference to the CSMF Accuracy. GPT-4 had the highest PCCC across all three age groups at 0.71, 0.73, and 0.61 for adult, child, and neonatal age groups. GPT-3.5 had the next highest PCCC for the adult and child age groups at 0.62 and 0.69 PCCC respectively, performing slightly worse than GPT-4 by 0.09 and 0.04 PCCC. InSilicoVA had the second highest PCCC for the neonatal age group at 0.56, which was 0.05 PCCC worse than GPT-4. Both InterVA-5 and InSilicoVA had the lowest PCCC for the adult and child age groups at 0.41 and 0.45 PCCC respectively, while InterVA-5 and GPT-3.5 had the lowest PCCC for the neonatal age group at 0.49 and 0.51 respectively. The PCCC for models had larger differences between the minimum and maximum PCCC for the adult and child age groups at 0.3 and 0.28 respectively than the neonatal age group at 0.12. For reference, all records were also coded by GPT-3.5, GPT-4, InterVA-5, and InSilicoVA for comparison to physician agreed records. For physician agreed records, models had 0.05 to 0.1 higher PCCC values and 0.01 to 0.05 higher CSMF accuracy values (see Figure \ref{fig-perf-allvsagree} in Appendix \ref{app-perf-misc}).

\begin{table}[h]
\caption{Overall model performance for physician agreed records (n=6939)}\label{tab-perf-overall}%
\begin{tabular}{@{}llll@{}}
\toprule
Model & PCCC  & CSMF Accuracy & PCCC Improvement\\
\midrule
GPT-4 & 0.61 & 0.79 & $\uparrow$ 0.03\\
GPT-3.5 & 0.58 & 0.74 & $\uparrow$ 0.14\\
InSilicoVA & 0.44 & 0.74 & $\uparrow$ 0.01\\
InterVA-5 & 0.43 & 0.78 & -\\
\botrule
\end{tabular}
\footnotetext{Note: Models sorted from highest to lowest PCCC. Improvements are from model below.}
\end{table}

>>>>

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
\usepackage{lstautogobble}%
%%%%

\lstset{
    texcl=true,
    basicstyle=\small\sf,
    commentstyle=\small\rm,
    mathescape=true,
    escapeinside={(*}{*)},
    breaklines=true,
    breakindent=0.5pt
}

@@

Recently, Large Language Models (LLM), leveraging massive datasets and deep learning approaches, have made advances in performing a variety of Natural Language Processing (NLP) tasks using free-text, such as question answering, code generation, and even medical diagnosis \cite{changSurveyEvaluationLarge2023,lundChattingChatGPTHow2023,svyatkovskiyIntelliCodeComposeCode2020,hauptAIGeneratedMedicalAdvice2023}. On November 30, 2022, a widely-available LLM called ChatGPT was released by OpenAI with capabilities of answering natural language text inquiries using training data up to September 2021. ChatGPT-3 was based on several Generative Pre-trained Transformer (GPT) models between 2018 to 2020, namely GPT-1 to GPT-3, which had notable differences in training data sizes of 5 gigabytes to 45 terabytes from web sources that resulted in 117 million to 175 billion parameter models \cite{wuBriefOverviewChatGPT2023}. On March 14, 2023, ChatGPT-4 was released with human-level performance on various professional and academic exams and benchmarks that outperformed ChatGPT-3 \cite{openaiGPT4TechnicalReport2023}. Given the limited usage of free-text open narratives in computer models for determining CODs, and recent advances in LLMs that leverage natural language text prompts, we conduct a case study with Sierra Leone deaths from VA in 2019 to 2022 to compare four models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, to physicians for determining CODs.

@@

Four computer models were used to assign COD for each of the 6939 physician agreed records: GPT-3.5, GPT-4, InterVA-5, and InSilicoVA. Each model required pre-processing of the 6939 records into input data, and standardization of output COD codes from models for performance evaluation as not all models produced comparable codes across outputs. Although each model can assign multiple CODs per record, only the first generated COD response from GPT-3.5 and GPT-4, and the most probable COD from InterVA-5 and InSilicoVA were used for evaluation. All model outputs were converted to CGHR-10 codes to evaluate performances of models for COD assignment relative to physicians.

@@

GPT-3.5 \cite{brownLanguageModelsAre2020} and GPT-4 \cite{openaiGPT4TechnicalReport2023} are LLMs that utilize deep neural networks with transformer architectures \cite{vaswaniAttentionAllYou2017} and reinforcement learning from human feedback \cite{ouyangTrainingLanguageModels2022,christianoDeepReinforcementLearning2017,stiennonLearningSummarizeHuman2020,wirthSurveyPreferencebasedReinforcement2017} to follow instructions from prompts and provide human-level responses, with known differences in GPT-4 possessing multimodal capabilites (e.g. image/voice input/output), more recent training data, and improved responses compared to ChatGPT-3 \cite{wuBriefOverviewChatGPT2023}. GPT-3.5 and GPT-4 were instructed with text prompts to assign CODs based on the open narrative. InterVA-5 and InSilicoVA are widely used and studied standard statistical models \cite{murrayUsingVerbalAutopsy2014,otiVerbalAutopsyInterpretation2010,benaraEvaluationMethodsAssigning2023,chandramohanEstimatingCausesDeath2021,tungaVerbalAutopsyModels2021,leitaoComparisonPhysiciancertifiedVerbal2014,jhaAutomatedPhysicianAssignment2019} for COD assignment in VAs under the openVA framework \cite{liOpenVAToolkitVerbal2023}. InterVA-5 applies Bayesian probabilistic modelling \cite{bayesEssaySolvingProblem1958} using a set of standardized symptoms from reports and related conditional probabilities from medical experts to assign CODs based on the highest probability \cite{byassIntegratedApproachProcessing2019,byassStrengtheningStandardisedInterpretation2012}. InSilicoVA improves upon InterVA (e.g. comparable probabilities across individuals, measures of uncertainty, and inclusion of additional data sources) with a hierarchical Bayesian framework and Markov Chain Monte Carlo (MCMC) simulations \cite{brooksMarkovChainMonte1998,chibMarkovChainMonte2001,hanMarkovChainMonte2001} to incorporate multiple sources of uncertainty for assigning CODs based on the highest probability \cite{mccormickProbabilisticCauseofDeathAssignment2016b}. For assigning CODs, GPT-3.5 and GPT-4 require prompts containing conversation-like textual instructions as input, while InterVA-5 and InSilicoVA require structured symptom and associated sociodemographic data as input.

@@

For GPT-3.5 and GPT-4, 6939 text prompts were generated for each physician agreed record as input to instruct the models to assign CODs based on the open narratives. Two types of text prompts were used: user prompts and system prompts. System prompts contained textual instructions to assign the role of a physician ICD-10 coder with expertise in Sierra Leone. The following system prompt was used for each record:

\begin{lstlisting}
You are a physician with expertisein determining underlying causes of death in Sierra Leone by assigning the most probable ICD-10 code for each death using verbal autopsy narratives. Return only the ICD-10 code without description. E.g. A00. If there are multiple ICD-10 codes, show one code per line.
\end{lstlisting}

\noindent User prompts contained textual instructions to perform coding of VA records based on the age, sex, and narrative of the deceased. The following template was used to generate user prompts for each record, where \verb|<age>| and \verb|<sex>| from the questionnaire, and \verb|<narrative>| from the narratives, were replaced with values from the data:

@@

The GPT-3.5 and GPT-4 Application Programming Interface (API) was accessed using Python version 3.11.4 and used to assign CODs for each record. GPT-3.5 used the \verb|gpt-3.5-turbo| model, while GPT-4 used the \verb|gpt-4-0613| model. The parameter \verb|temperature| for GPT-3.5 and GPT-4, representing the sampling temperature ranging from 0 to 2 (default of 1), was set to 0 to produce more deterministic outputs \cite{openaiOpenAIPlatformAPI2024}. Higher values closer to 2 may produce less deterministic outputs, while lower values closer to 0 produce more deterministic outputs.

@@

Of the 6939 input records, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA were able to assign CODs for 6939 (100\%), 6935 (\textgreater 99\%), 6830 (98\%), 6830 (98\%) records respectively. All 6830 (100\%) InterVA-5 and InSilicoVA records with WHO VA 2016 v1.5 output codes \cite{worldhealthorganizationVerbalAutopsyStandards2016} were converted into ICD-10 codes respectively. After all model outputs were converted to ICD-10 codes, they were further converted to CGHR-10 codes. The 6939 GPT-3.5 and 6935 GPT-4 output records with ICD-10 codes were converted into 6930 (\textgreater 99\%) and 6931 (\textgreater 99) records with CGHR-10 codes, where \textless 1\% (9 and 8) records did not have matching CGHR-10 codes respectively. The 6830 InterVA-5 and InSilicoVA records with ICD-10 codes were converted into 6802 (\textgreater 99\%) and 6726 (98\%) records with CGHR-10 codes respectively, where 28 (\textless 1\%) and 104 (1\%) of records could not be converted into CGHR-10 codes.

@@

Of the 6939 records, GPT-4 had the highest PCCC followed by GPT-3.5, InSilicoVA, and InterVA-5 at 0.61, 0.58, 0.44, and 0.43 PCCC respectively (Table \ref{tab-perf-overall}). GPT-3.5 and GPT-4 had improvements ranging from 0.14-0.18 PCCC over InSilicoVA and InterVA-5, while GPT-4 slightly improved over GPT-3.5 by 0.03 PCCC. CSMF accuracies were above 0.7 for all models at 0.79, 0.74, 0.74, and 0.78 for GPT-4, GPT-3.5, InSilicoVA, and InterVA-5 respectively. Figure \ref{fig-perf-agegroup} shows the PCCC performance across age groups with reference to the CSMF Accuracy. GPT-4 had the highest PCCC at 0.65 and 0.62 for the adult and neonatal age groups, while GPT-3.5 had the highest PCCC at 0.57 for the child age group with GPT-4 slightly below it at 0.54. GPT-3.5 had the second highest PCCC at 0.56 fo the adult age group, while InSilicoVA had the second highest PCCC for the neonatal age group at 0.56. Both InterVA-5 and InSilicoVA had the lowest PCCC for the adult and child age groups at 0.41 and 0.45 PCCC respectively, while InterVA-5 and GPT-3.5 had the lowest PCCC for the neonatal age group at 0.49 and 0.51 respectively. The PCCC for models had larger differences between the minimum and maximum PCCC for the adult and child age groups at 0.3 and 0.28 respectively than the neonatal age group at 0.12. For reference, all records were also coded by GPT-3.5, GPT-4, InterVA-5, and InSilicoVA for comparison to physician agreed records. For physician agreed records, models had 0.05 to 0.1 higher PCCC values and 0.01 to 0.05 higher CSMF accuracy values (see Figure \ref{fig-perf-allvsagree} in Appendix \ref{app-perf-misc}).

\begin{table}[h]
\caption{Overall model performance for physician agreed records (n=6939)}\label{tab-perf-overall}%
\begin{tabular}{@{}llll@{}}
\toprule
Model & PCCC  & CSMF Accuracy & PCCC Improvement\\
\midrule
GPT-4 & 0.61 & 0.79 & $\uparrow$ 0.03\\
GPT-3.5 & 0.58 & 0.74 & $\uparrow$ 0.14\\
InSilicoVA & 0.44 & 0.74 & $\uparrow$ 0.01\\
InterVA-5 & 0.43 & 0.78 & -\\
\botrule
\end{tabular}
\footnotetext{Note: Models sorted from highest to lowest PCCC. PCCC improvements are from model below.}
\end{table}
